Accurate binding of sodium and calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-11-07 Manuscript/manuscript.tex
Removing TODOs: acknowledgments, SI figures, additional references, temperatures in captions etc.

\subsection{Electronic continuum correction for lipid bilayers}\label{section:ecc}
The lack of electronic polarizability in standard MD simulation
force fields has been considered a serious issue since the early days of
lipid bilayer simulations.
In this work, we circumvent the rather demanding explicit inclusion
of electronic polarization effects \cite{lucas12,chowdhary13} by
including the electronic part of polarizability in lipid
bilayer simulations implicitly via 
the electronic continuum correction (ECC)~\cite{leontyev11}. 
Technically, this is similar to the phenomenological charge-scaling applied in 
earlier studies of aqueous solutions or ionic liquids \cite{jonsson86,egberts94}.
\todo{We should also cite papers where empirical scaling was used ionic liquids DOI: 10.1002/anie.201308760}
However, the present concept of ECC is
physically well justified and rigorously derived~\cite{leontyev09, leontyev10, leontyev11, leontyev14}.

@@

While the scaling factor of $f_q = 0.75$ for ions in water improves their description and is physically well justified within the ECC theory \cite{Leontyev2011},
it is not a priori  clear whether the same factor should be used for partial charges in molecules,
e.g., lipids in our case. Unlike the total charge of an ion, atomic partial charges within molecules are not physical observables.
Several schemes exist for the assignment of partial charges for biomolecules~\cite{Hu2007}.
Currently, the most commonly employed scheme is the restrained electrostatic potential (RESP) method~\cite{RESP_paper,Singh1984}.
By construction, partial charges currently implemented in force fields may already 
include  to some extend the electronic polarizability effects of the solvent, i.g.,
the RESP charges may be subsequently modified to fit certain experimental observables \todo{This needs a citation - Joe please pick 1-2 papers}. 
Thus, a consistent application of the ECC scaling factor, $f_q$, to the partial charges in molecules as
included in the available force fields does not necessarily have to follow the above relation $f_q = \epsilon _{el} ^{-1/2}$. 
Instead, a consistent scaling factor would lie between the value of 0.75 i.e., (no electronic polarizability included in the original partial charges)
and $1$ (i.e., electronic polarizability fully included in the original partial charges).
It is important to note that even though the lipid bilayer interface has a large gradient in
the total dielectric constant, the electronic part of polarizability can be considered to be almost constant
as measurements give values around 2 
for almost any biologically relevant systems \cite{some_original_work, leontyev11}.

@@

\begin{figure}[tb!]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/PN_angle_OrdPars-A-B_L14-ECCL17_q80_sig89_surf.pdf}
  \caption{\label{OrderParameterCHANGESsurf}
    The changes of headgroup order parameters and P-N vector orientation as a function of
    cationic surfactant (dihexadecyldimethylammonium) in POPC bilayer from simulations
    and experiments \cite{scherer89}.
  }
\end{figure}

@@

\begin{figure}[htb!]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/PN_angle_OrdPars-A-B-g3_L14-ECCL17_q80_sig89_CaCl.pdf}
  \caption{\label{fig:delta_ordPar_CaCl}
    Changes of the head group order parameters of POPC bilayer as functions of the CaCl$_2$ concentration
    are shown from simulations with different force fields together with experimental data 
    (DPPC \cite{akutsu81} and POPC \cite{altenbach84}). 
    Ion concentrations in bulk water are shown in x-axis. 
    Values from simulations are calculated from the of cation number density $C_{np}$
    from the region at the simulatin box edge with the constant ion concentration as [ion]=$C_{np}/0.602$.
    Simulation data with Lipid14 and \AA{}qvist ion parameters are taken directly from
    Refs.~\cite{lipid14POPC0mMNaClfiles,lipid14POPC350mMCaClfiles,lipid14POPC350mMCaClfilesNC}.
  }
  \todo{Add temperatures for POPC and DPPC experiments} 
  %\todo{Experimental values from \cite{akutsu81} to be put in the $g_3$ figure:
    %the value of $g_3$ order parameter of DPPC was -0.214 
    %in the absence of ions [average of two closely spaced splittings] and -0.211 in the presence
    %of 0.35 M CaCl$_2$ (at 59 Celcius). No effect of ions could be detected on DPPC bilayers labeled
    %at the C-2 segments of both fatty acyl chains.'' }
\end{figure}

\begin{figure}[htb!]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/OrdPars-A-B_L14-ECCL17_q80_sig89_NaCl.pdf}
  \caption{\label{fig:delta_ordPar_NaCl}
    The changes of the head group order parameters of POPC bilayer as a function of NaCl concentrations
    from simulations with different force fields together with experimental data for DPPC~\cite{akutsu81} and POPC~\cite{altenbach84}.
    Ion concentrations in x-axis are calculated as in Fig.~\ref{fig:delta_ordPar_CaCl}.
    Simulation data with Lipid14 and \AA{}qvist ion parameters is taken directly from
    Refs.~\cite{lipid14POPC0mMNaClfiles,lipid14POPC1000mMNaClfiles}.
  }
  %\todo{Add also the POPC result to the figure. Ref.\cite{altenbach84} gives for POPC at 1M of NaCl $\Delta S^\alpha=(6.1-5.5)*0.00784=0.004704$.} \\
  \todo{Add temperatures to the experimental data.} 
\end{figure}

@@

% If you have acknowledgments, this puts in the proper section head.
\begin{acknowledgments}
% Put your acknowledgments here.
P.J. acknowledges support from the Czech Science Foundation (grant no. 16-01074S) 
and 600 from the Academy of Finland via the FiDiPro award.
\todo{Add acknowledgments of Jiri Kolafa.}
\todo{Joe: Thank Metacentrum.}
\end{acknowledgments}

@@

>>>>

\subsection{Electronic continuum correction for lipid bilayers}\label{section:ecc}
The lack of electronic polarizability in standard MD simulation
force fields has been considered a serious issue since the early days of
lipid bilayer simulations.
In this work, we circumvent the rather demanding explicit inclusion
of electronic polarization effects \cite{lucas12,chowdhary13} by
including the electronic part of polarizability in lipid
bilayer simulations implicitly via 
the electronic continuum correction (ECC)~\cite{leontyev11}. 
Technically, this is similar to the phenomenological charge-scaling applied in 
earlier studies of aqueous solutions or ionic liquids \cite{jonsson86,egberts94,beichel14}.
However, the present concept of ECC is
physically well justified and rigorously derived~\cite{leontyev09, leontyev10, leontyev11, leontyev14}.

@@

While the scaling factor of $f_q = 0.75$ for ions in water improves their description and is physically well justified within the ECC theory \cite{Leontyev2011},
it is not a priori  clear whether the same factor should be used for partial charges in molecules,
e.g., lipids in our case. Unlike the total charge of an ion, atomic partial charges within molecules are not physical observables.
Several schemes exist for the assignment of partial charges for biomolecules~\cite{Hu2007}.
Currently, the most commonly employed scheme is the restrained electrostatic potential (RESP) method~\cite{RESP_paper,Singh1984}.
By construction, partial charges currently implemented in force fields may already 
include  to some extent the electronic polarizability effects of the solvent, i.g.,
the RESP charges may be modified during fitting or subsequently after it 
to improve certain experimental observables 
\cite{RESP_paper,Singh1984,jorgensen96,ipolq2013,benavides17}. 
Thus, a consistent application of the ECC scaling factor, $f_q$, to the partial charges in molecules as
included in the available force fields does not necessarily have to follow the above relation $f_q = \epsilon _{el} ^{-1/2}$. 
Instead, a consistent scaling factor would lie between the value of 0.75 i.e., (no electronic polarizability included in the original partial charges)
and $1$ (i.e., electronic polarizability fully included in the original partial charges).
It is important to note that even though the lipid bilayer interface has a large gradient in
the total dielectric constant, the electronic part of polarizability can be considered to be almost constant
as measurements give values around 2 
for almost any biologically relevant systems \cite{some_original_work, leontyev11}.

@@

\begin{figure}[tb!]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/PN_angle_OrdPars-A-B_L14-ECCL17_q80_sig89_surf.pdf}
  \caption{\label{OrderParameterCHANGESsurf}
    The changes of headgroup order parameters and P-N vector orientation as a function of
    a molar fraction of the cationic surfactant dihexadecyldimethylammonium in a POPC bilayer
    from simulations and experiments \cite{scherer89}.
  }
\end{figure}

@@

\begin{figure}[htb!]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/PN_angle_OrdPars-A-B-g3_L14-ECCL17_q80_sig89_CaCl.pdf}
  \caption{\label{fig:delta_ordPar_CaCl}
    Changes of the head group order parameters and P-N vector orientation of a POPC bilayer 
    as a function of the CaCl$_2$ concentration
    are shown from simulations with different force fields together with experimental data 
    (DPPC (323\,K) \cite{akutsu81} and POPC (313\,K) \cite{altenbach84}). 
    Ion concentrations in bulk water are shown in x-axis. 
    Bulk concentrations from simulations are calculated 
    from the farthest point from the lipid bilayer in the aqueous phase
    with an error estimate of 10\,mM.
    The error estimate for the simulation with Lipid14 and ECC-ions is an order of magnitude larger, 
    because the simulation cell was too small (shown if Fig.~\ref{fig:cacl-dens}). 
    Simulation data with Lipid14 and \AA{}qvist ion parameters are taken directly from
    Refs.~\cite{lipid14POPC0mMNaClfiles,lipid14POPC350mMCaClfiles,lipid14POPC350mMCaClfilesNC}.
  }
  %\todo{Experimental values from \cite{akutsu81} to be put in the $g_3$ figure:
    %the value of $g_3$ order parameter of DPPC was -0.214 
    %in the absence of ions [average of two closely spaced splittings] and -0.211 in the presence
    %of 0.35 M CaCl$_2$ (at 59 Celcius). No effect of ions could be detected on DPPC bilayers labeled
    %at the C-2 segments of both fatty acyl chains.'' }
\end{figure}

\begin{figure}[htb!]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/OrdPars-A-B_L14-ECCL17_q80_sig89_NaCl.pdf}
  \caption{\label{fig:delta_ordPar_NaCl}
    The changes of the head group order parameters of a POPC bilayer as a function of NaCl concentration
    from simulations with different force fields together with 
    experimental data for DPPC (323\,K) \cite{akutsu81} and POPC (313\,K) \cite{altenbach84}.
    Ion concentrations in x-axis are calculated as in Fig.~\ref{fig:delta_ordPar_CaCl}.
    Simulation data with Lipid14 and \AA{}qvist ion parameters are taken directly from
    Refs.~\cite{lipid14POPC0mMNaClfiles,lipid14POPC1000mMNaClfiles}.
  }
  %\todo{Add also the POPC result to the figure. Ref.\cite{altenbach84} gives for POPC at 1M of NaCl $\Delta S^\alpha=(6.1-5.5)*0.00784=0.004704$.} \\
\end{figure}

@@

% If you have acknowledgments, this puts in the proper section head.
\begin{acknowledgments}
% Put your acknowledgments here.
P.J. acknowledges support from the Czech Science Foundation (grant no. 16-01074S) 
and 600 from the Academy of Finland via the FiDiPro award.
J.K. acknowledges support from the Czech Science Foundation, Project No. 15-12386S.
Computational resources were supplied by the Ministry of Education, Youth and Sports of the Czech Republic under the Projects CESNET (Project No. LM2015042) and CERIT-Scientific Cloud (Project No. LM2015085) provided within the program Projects of Large Research, Development and Innovations Infrastructures.
\end{acknowledgments}

@@

\begin{figure}[tbp!]
  \centering
  \includegraphics[width=16.0cm]{../Fig/ipython_nb/Order-parameters_exp-L14-ECCL17_q80_sig89_GMX-oMM_compar.pdf}
  \caption{\label{fig:ordPars_actual_GMX_oMM_compar}
    Order parameters of POPC head group, glycerol backbone and acyl chains 
    from simulations with the Lipid14 \cite{dickson14} and the ECC-lipids models
    simulated with GROMACS~5.1.4 \cite{Abraham15} and openMM~7 \cite{openmm7} 
    compared with the experimental values from \cite{ferreira13}.
    The size of the markers for the head group order parameters correspond to
    the error estimate $\pm 0.02$ for experiments \cite{botan15,ollila16},
    while the error estimate for simulations is $\pm 0.005$.
    The size of the points for acyl chains are decreased by a factor of 3 to improve the clarity of the plot.
  }
\end{figure}

\begin{figure}[tbp!]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/PN_angle_OrdPars-A-B-g3_L14-ECCL17_q80_sig89_surf_GMX-oMM_compar.pdf}
  \caption{\label{fig:ordPars_surf_GMX_oMM_compar}
    The changes of headgroup order parameters and P-N vector orientation as a function of
    a molar fraction of the cationic surfactant dihexadecyldimethylammonium in a POPC bilayer
    from simulations with the ECC-lipids models
    simulated with GROMACS~5.1.4 \cite{Abraham15} and openMM~7 \cite{openmm7} 
    compared with the experimental values from \cite{scherer89}.
    The size of the markers for the head group order parameters correspond to
    the error estimate $\pm 0.02$ for experiments \cite{botan15,ollila16},
    while the error estimate for simulations is $\pm 0.005$.
  }
\end{figure}

\begin{figure}[tbp!]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/PN_angle_OrdPars-A-B-g3_L14-ECCL17_q80_sig89_CaCl_GMX-oMM_compar.pdf}
  \caption{\label{fig:ordPars_cacl_GMX_oMM_compar}
    Changes of the head group order parameters and P-N vector orientation of a POPC bilayer 
    as a function of the CaCl$_2$ concentration
    from simulations with the Lipid14 \cite{dickson14} and the ECC-lipids models
    simulated with GROMACS~5.1.4 \cite{Abraham15} and openMM~7 \cite{openmm7} 
    (DPPC (323\,K) \cite{akutsu81} and POPC (313\,K) \cite{altenbach84}). 
    Ion concentrations in bulk water are shown in x-axis. 
    Bulk concentrations from simulations are calculated 
    from the farthest point from the lipid bilayer in the aqueous phase
    with an error estimate of 10\,mM.
  }
\end{figure}
